---
title: "Tempus Data Analysis 5/23/2019"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    df_print: paged
    toc: true
    theme: spacelab
    highlight: textmate
---
#Data structure and format
```{r Database setup, message=FALSE, echo=FALSE, results='asis', warning=FALSE, echo=FALSE, include=FALSE}
knitr::opts_chunk$set
knitr::kable

setwd("~/Dropbox/UCOncology/Research/Abstracts/Tempus/Tempus-Analysis")
library(tidyr)
library(dplyr)
library(knitr)
library(ggplot2)
library(survival)
library(survminer)
library(stringr)
tempusclinical <- as.data.frame(read.csv(file="~/Research-Local/Tempus-Local/Input/ChicagoMultiethnic_clinical_2019Feb05.csv", header=TRUE, sep=",")) #127 patients
tempusmutation <- as.data.frame(read.csv(file="~/Research-Local/Tempus-Local/Input/TempusReportDataset-2019Feb1_Final_mutation.csv", header=TRUE, sep=","))
tempusreports <- as.data.frame(read.csv(file="~/Research-Local/Tempus-Local/Input/TempusReportDataset-2019Feb1_Final_report.csv", header=TRUE, sep=",")) #139 reports = 8 lymph nodes + 4 patients without clinical information
tempusimmune <- as.data.frame(read.csv(file="~/Research-Local/Tempus-Local/Input/Immune_data_v1-05132019.csv", header=TRUE, sep=",")) #74 patients with immune profiles calculated 
tempusmutation <- tempusmutation %>% group_by(StudyID, AccessionNo) %>% mutate(genecount = n_distinct(Gene))
tempusmutation <- tempusmutation %>% group_by(StudyID, AccessionNo, Gene) %>%mutate(variantcount = seq_along(Gene))
tempusmutationtemp <- tempusmutation # saving the original mutation data and using a copy table to make modifications
```

```{r Formatting the mutation data for merging}
tempusmutationtemp$GeneInfo <- paste (tempusmutationtemp$variantcount, tempusmutationtemp$variant_name_C, tempusmutationtemp$variant_name_P, tempusmutationtemp$NM_number, tempusmutationtemp$typeofvariant, tempusmutationtemp$VAF, tempusmutationtemp$COSMICCases, sep='\t') # combines variant-specific variables into a temporary column for now

tempusmutationtemp <- tempusmutationtemp %>% group_by(StudyID, AccessionNo) %>%mutate(maxvariantcount = max(variantcount))

tempusmutationtemp <- dplyr::select(tempusmutationtemp, -c(biologicalmeaning, variant_name_C, variant_name_P, NM_number, typeofvariant, VAF, COSMICCases, variantcount, Notes)) #drops excess columns for merge purposes

tempusmutationtemp <- tempusmutationtemp %>% spread(Gene,GeneInfo) #Creates individual columns for each gene with the Gene Information as the value, every variant has its own row

tempusmutationtemp$V1=NULL #Clears artificial columns that were created in the process of formatting the dataset
tempusmutationtemp$X=NULL

tempusmutationtemp <- tempusmutationtemp %>% group_by(StudyID, AccessionNo, PanelName, Normalspecimen, Normalcollecteddate, Normalreceiveddate, Tumorspecimen, Tumorcollecteddate, Tumorreceiveddate, TumorPercentage, TumorMutationalBurden, PercentileTumorMutationalBurd, MicrosatelliteInstabilityStatu, DateSigned, genecount, maxvariantcount) %>% summarise_all(funs(toString(na.omit(.), sep='\t'))) #This flattens all of the rows -> RIGHT NOW each patient is on one line and multiple mutations for the same gene are in one entry as are their variant information for ease of merge

tempusmerge <- merge (tempusclinical, tempusmutationtemp, by="StudyID", all.x=TRUE, all.y=FALSE, sort=FALSE) #Merging the clinical and mutational data into one database, reports are excluded if they do not have clinical information

tempusimmune$StudyID <- tempusimmune$study_id
tempusimmunetemp <- dplyr::select(tempusimmune, -c(order_id, TMB, msi_status, study_id)) #drops excess columns for merge purposes
tempusimmunetemp <- filter (tempusimmunetemp, !(is.na(b))) #Removes rows with no data (as some of these are duplicates and confuse the merge process)
tempusimmunetemp <- filter (tempusimmunetemp, (StudyID != "PN0198" & StudyID != "PN0206")) #Removes 2 duplicated rows in which the reason for duplication is not completely clear; this will need to be clarified later.

tempusmerge <- merge (tempusmerge, tempusimmunetemp, by="StudyID", all.x=TRUE) #Merging the immune information if available with the existing dataset. Data was excluded if clinical and mutational information were not also available. 
```

#Data cleaning/new variable generation
```{r Data cleaning}
tempusmerge$age50 <- ifelse(tempusmerge$age_at_diagnosis < 50, 1, 0) #binary variable for age < 50 (which equals 1)
tempusmerge$neoadj <- ifelse (((as.Date(as.character(tempusmerge$date_surg1), format="%m/%d/%Y")-as.Date(as.character(tempusmerge$date_cx1), format="%m/%d/%Y"))>0) & (tempusmerge$Stage!="4"), 1, 0) #determining if pts had neoadjuvant chemotherapy
tempusmerge$HRstatus <- ifelse (tempusmerge$ER=="Pos" | tempusmerge$PR=="Pos", 1, 0) #Hormone-receptor positive status

tempusmerge$subtype <- ifelse (tempusmerge$HER2.x=="Neg" & tempusmerge$HRstatus==0, "TNBC", NA) #Characterizing the subtype variable
tempusmerge$subtype <- ifelse (tempusmerge$HER2.x=="Pos" & tempusmerge$HRstatus==0, "HER+/HR-", tempusmerge$subtype) #Characterizing the subtype variable
tempusmerge$subtype <- ifelse (tempusmerge$HER2.x=="Pos" & tempusmerge$HRstatus==1, "HER+/HR+", tempusmerge$subtype) #Characterizing the subtype variable
tempusmerge$subtype <- ifelse (tempusmerge$HER2.x=="Neg" & tempusmerge$HRstatus==1, "HER-/HR+", tempusmerge$subtype) #Characterizing the subtype variable

tempusmerge$bw <- ifelse (tempusmerge$race=="Black", "Black", NA) #Characterizing black/white alone
tempusmerge$bw <- ifelse (tempusmerge$race=="White", "White", tempusmerge$bw) 
tempusmerge$bw <- ifelse (tempusmerge$hispanic=="Yes", NA, tempusmerge$bw) 

tempusmerge$vitalstatus <- ifelse (tempusmerge$vitalstatus=="0 Dead", "0", "1") #Factorizing vital status

tempusmerge$vitalstatussurv <- tempusmerge$vitalstatus 

tempusmerge$vitalstatussurv <- ifelse (tempusmerge$vitalstatussurv=="1", 2, tempusmerge$vitalstatussurv)

tempusmerge$vitalstatussurv <- ifelse (tempusmerge$vitalstatussurv=="0", 1, tempusmerge$vitalstatussurv)

tempusmerge$vitalstatussurv <- ifelse (tempusmerge$vitalstatussurv=="2", 0, tempusmerge$vitalstatussurv) #switching 0 and 1 so they accurately reflect survival failure

tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "1A",1,NA)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "1B",1, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "1",1, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "2",2, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "2A",2, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "2B",2, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "3",3, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "3A",3, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "3B",3, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "3C",3, tempusmerge$stage14)
tempusmerge$stage14 <- ifelse (tempusmerge$Stage %in% "4",4, tempusmerge$stage14) #Numerical stage condensing

tempusmerge$nodeposneg <- ifelse (tempusmerge$RegionalNodePositive>0, 1, 0) #At least one positive lymph node

tempusmerge$recurrenceposneg <- ifelse (tempusmerge$type1strecurrence=="00 Patient became disease-free after treatment and has not had a recurrence.", 0, 1) #Positive/negative recurrence status alone without differentiation of site. 
tempusmerge$recurrenceposneg <- ifelse (tempusmerge$type1strecurrence=="99 It is unknown whether the disease has recurred or if the patient was ever disease-free", NA, tempusmerge$recurrenceposneg) #Cleaning unknown recurrence variables from this
tempusmerge$recurrenceposneg <- ifelse (tempusmerge$type1strecurrence=="70 Since diagnosis, patient has never been disease-free. This includes cases with distant metastasis at diagnosis, systemic disease, unknown primary, or minimal disease that is not treated", NA, tempusmerge$recurrenceposneg) #Cleaning unknown recurrence variables from this

tempusmerge$TMB <- tempusmerge$TumorMutationalBurden
tempusmerge$TMB <- str_replace(tempusmerge$TMB, pattern = " m/MB", replacement = "")
tempusmerge$TMB <- as.numeric(tempusmerge$TMB) #Numeric interpretation of mutational burden

tempusmerge$timefu <- ifelse (tempusmerge$neoadj==1, ((as.Date(as.character(tempusmerge$date_FU), format="%m/%d/%Y"))-(as.Date(as.character(tempusmerge$date_cx1), format="%m/%d/%Y")))/360, ((as.Date(as.character(tempusmerge$date_FU), format="%m/%d/%Y"))-(as.Date(as.character(tempusmerge$date_surg1), format="%m/%d/%Y")))/360)
tempusmerge$timefu <- ifelse(is.na(tempusmerge$timefu), ((as.Date(as.character(tempusmerge$date_FU), format="%m/%d/%Y")-as.Date(as.character(tempusmerge$date_surg1), format="%m/%d/%Y")))/360, tempusmerge$timefu)
tempusmerge$timefu <- ifelse(tempusmerge$timefu<0, NA, tempusmerge$timefu) #duration of follow up for survival curves in years

tempusmergedupclean <- tempusmerge
tempusmergedupclean <- tempusmergedupclean[with(tempusmergedupclean, ave(TumorPercentage, StudyID, FUN=max)==TumorPercentage),] #keep duplicates by higher %
tempusmergedupclean <- tempusmergedupclean[with(tempusmergedupclean, ave(TMB, StudyID, FUN=max)==TMB),] #keep duplicates by higher TMB if same source, %
```

#Data summary
We have clinical information for `r length(unique(tempusclinical$StudyID))` patients and mutation information for `r length(unique(tempusreports$StudyID))` patients. We have immune information for `r length(unique(tempusimmunetemp$StudyID))` patients.

We are missing clinical information for the following 4 patients. These are not included in the merged data set. 
```{r Clinical missing, echo=FALSE}
setdiff(tempusreports$StudyID, tempusclinical$StudyID)
```

```{r Mutiple samples present}
duplicates <- tempusreports %>% filter(duplicated(StudyID) | duplicated(StudyID, fromLast = TRUE)) %>% select(StudyID, AccessionNo, Tumorspecimen, TumorPercentage)
duplicates
```
`r length(unique(duplicates$StudyID))` patients have duplicate reports (most often primary tumor and lymph node) submitted with the above characteristics. These patients are included in the merged data set but in these cases, for any analysis, we counted each patient individually/once using the specimen that offered the best tumor percentage as relevant. 

Of note, among the immune reports, IDs PN0198 (likely due to breast and lymph node samples) and PN0206 (unclear reasons) had 2 separate reports with differing values. As such, these immune reports were excluded from the analysis.  

```{r Lymph nodes only present}
lymphnodes <- tempusreports %>% filter((Tumorspecimen=="Lymph node") & !(duplicated(StudyID) | duplicated(StudyID, fromLast = TRUE))) %>% select(StudyID, AccessionNo, Tumorspecimen, TumorPercentage)
lymphnodes
```
`r length(unique(lymphnodes$StudyID))` patients have only lymph node submitted. These patients are included in the merged data set and treated as though they had primary tumor (as this was their only tumor specimen). 

#Clinical summary
Low-yield variables: Date and age of enrollment (38 patients), Sex (1 male patient), Insurance (1 patient with unknown insurance, 126 patients insured), Alcohol history (90 patients not reported), gravidity/parity (26 patients reported), menopause (26 patients reported), distant metastatic sites (only 2 sites reported), histology (14 patients with at least partial lobular disease), surgical margin (4 patients with ? residual disease), 2nd primary (only one patient reported), ki67: not known in 116/127 patients

Variables not incorporated for this analysis: Marital status at diagnosis, breast primary site, laterality, TNM edition and clinical staging (Rather than pathologic staging), lymphovascularinvasion (64 patients only reported), radiation variables

```{r Clinical summary, echo=FALSE}
hist(tempusclinical$age_at_diagnosis)
```
This is the age distribution of patients at time of diagnosis. This population skews young but does have representation throughout the age range. 

```{r Ongoing clinical summary, echo=FALSE}
kable(table(tempusmergedupclean$age50), caption = "Distribution of patients younger than age 50 (with 0 > 50 and 1 < 50", col.names =c("age<50", "Freq"))
kable(table(tempusclinical$race, tempusclinical$hispanic), caption = "Self-reported races in our cohort (Y/N refers to Hispanic ethnicity)", col.names =c("Not-Hispanic", "Yes-Hispanic"))
kable(table(tempusclinical$tobacco_history), caption = "Tobacco exposure", col.names=c("Reported smoking history", "Freq"))
kable(table(tempusclinical$vitalstatus), caption = "Alive or dead status last known", col.names=c("Vital status", "Freq"))
kable(table(tempusmergedupclean$recurrenceposneg), caption = "Patients who experienced recurrence", col.names=c("Recurrence", "Freq"))
kable(table(tempusmergedupclean$nodeposneg), caption = "Fraction of patients with at least one node positive on study", col.names=c("1+ positive nodes", "Freq"))
kable(table(tempusclinical$grade), caption = "Grade information (known for 125/127 pts)", col.names=c("Grade", "Freq"))
kable(table(tempusclinical$Stage), caption = "Distribution of analytical stage based on a synthesis of clinical stage and pathological stage information", col.names=c("Compiled stage", "Freq"))
kable(table(tempusmergedupclean$HRstatus, tempusmergedupclean$HER2.x), caption="Hormone receptor status vs HER2 status. 0 = negative and 1=positive", col.names=c("Unk", "HER-neg", "HER2-pos"), row.names=TRUE)
kable(table(tempusmergedupclean$subtype), caption="Overall subtype distribution", col.names=c("Subtype", "Freq"))
kable(table(tempusclinical$chemotherapy), caption="Recieved any type of chemotherapy", col.names=c("Chemotherapy", "Freq"))
kable(table(tempusmergedupclean$neoadj), caption="Patients with neoadjuvant therapy", col.names=c("Neoadjuvant therapy", "Freq"))
kable(table(tempusclinical$radiotherapy), caption="Recieved any type of radiotherapy", col.names=c("Radiotherapy", "Freq"))
kable(table(tempusclinical$surgery1), caption="Initial surgery performed at time of diagnosis and initial treatment", col.names=c("Initial surgery type", "Freq"))
kable(table(tempusclinical$hormonetherapy), caption="Recieved any type of hormone therapy", col.names=c("Hormone therapy", "Freq"))
```
Of note, with mutation testing prior to sequencing, we had 1 patient each report a BRCA 1 and BRCA 2 mutation. 

#Mutation summary
Variables not incorporated for this analysis: Percentile mutational burden (incomplete)

We have representation of `r length(unique(tempusmutation$Gene))` genes with mutations in this data set (including both primary tumors and lymph nodes). The most common mutations (incorporating BOTH pathogenic + VUS) were as follows: 
```{r Mutation summary and tumor purity, echo=FALSE}

numGenes <- colSums(!((tempusmergedupclean[,105:663])==""))
numGenes["ERBB2/HER2"] <- sum(numGenes["ERBB2"], numGenes["ERBB2 (HER2)"], numGenes["ERBB2(HER2)"], numGenes["ERBB2HER2)"])

head(sort(numGenes,decreasing=TRUE), 10) #Frequency of gene mutations per unique patients 

hist(tempusmergedupclean$TumorPercentage, main="Distribution of purity of tumor samples")
```
This is the tumor purity distribution of patients at time of diagnosis. 30% purity is conisdered sufficient to perform MSI testing, for example. 104 of our 127 samples met this criteria. 

```{r Oncoprint}
tempusoncoprint <- tempusmutation %>% filter(AccessionNo %in% tempusmergedupclean$AccessionNo) %>% select(StudyID, AccessionNo, Gene, variant_name_P, typeofvariant)

tempusoncoprint$alteration <- NA
tempusoncoprint$variant <- NA

tempusoncoprint$variant <- ifelse(tempusoncoprint$typeofvariant=="copy number gain" | tempusoncoprint$typeofvariant=="Copy number gain" | tempusoncoprint$typeofvariant=="Copy Number gain" | tempusoncoprint$typeofvariant=="copy numbre gain" | tempusoncoprint$typeofvariant=="copy number loss" | tempusoncoprint$typeofvariant=="Copy number loss" | tempusoncoprint$typeofvariant=="Copy numberloss" | tempusoncoprint$typeofvariant=="Copy Number Loss", "CNA", tempusoncoprint$variant)

tempusoncoprint$alteration <- ifelse(tempusoncoprint$typeofvariant=="copy number gain" | tempusoncoprint$typeofvariant=="Copy number gain" | tempusoncoprint$typeofvariant=="Copy Number gain" | tempusoncoprint$typeofvariant=="copy numbre gain", "GAIN", tempusoncoprint$alteration)

tempusoncoprint$alteration <- ifelse (tempusoncoprint$typeofvariant=="copy number loss" | tempusoncoprint$typeofvariant=="Copy number loss" | tempusoncoprint$typeofvariant=="Copy numberloss" | tempusoncoprint$typeofvariant=="Copy Number Loss", "HETLOSS", tempusoncoprint$alteration)

tempusoncoprint$variant <- ifelse(tempusoncoprint$typeofvariant=="frameshift" | tempusoncoprint$typeofvariant=="Frameshift" | tempusoncoprint$typeofvariant=="Frameshift - LOF" | tempusoncoprint$typeofvariant=="Frameshift VAF:" | tempusoncoprint$typeofvariant=="Frameshift- LOF" | tempusoncoprint$typeofvariant=="Frameshift- LOF VAF" | tempusoncoprint$typeofvariant=="frameshift-LOF" | tempusoncoprint$typeofvariant=="FrameshiftVAF:" | tempusoncoprint$typeofvariant=="Inframe deletion" | tempusoncoprint$typeofvariant=="Inframe deletion - GOF" | tempusoncoprint$typeofvariant=="inframe insertion" | tempusoncoprint$typeofvariant=="Inframe insertion" | tempusoncoprint$typeofvariant=="Inframe insertion - GOF" | tempusoncoprint$typeofvariant=="rameshift" | tempusoncoprint$typeofvariant=="Frameshift LOF", "INFRAME", tempusoncoprint$variant)

tempusoncoprint$variant <- ifelse(tempusoncoprint$typeofvariant=="Missense  variant" | tempusoncoprint$typeofvariant=="Missense varian(exon 10) - GOF" | tempusoncoprint$typeofvariant=="missense variant" | tempusoncoprint$typeofvariant=="Missense variant" | tempusoncoprint$typeofvariant=="Missense Variant" | tempusoncoprint$typeofvariant=="Missense variant -" | tempusoncoprint$typeofvariant=="Missense variant - GOF" | tempusoncoprint$typeofvariant=="Missense variant - LOF:" | tempusoncoprint$typeofvariant=="Missense variant (exon 10) - GOF" | tempusoncoprint$typeofvariant=="Missense variant (exon 10)- GOF" | tempusoncoprint$typeofvariant=="Missense variant (exon 14) - GOF" | tempusoncoprint$typeofvariant=="Missense variant (exon 21) - GOF" | tempusoncoprint$typeofvariant=="Missense variant (exon 21) - GOF VAF" | tempusoncoprint$typeofvariant=="Missense variant (exon 5) - GOF" | tempusoncoprint$typeofvariant=="Missense Variant (exon 8)- GOF" | tempusoncoprint$typeofvariant=="Missense Variant (exon 8)- GOF" | tempusoncoprint$typeofvariant=="Missense variant 창\u0080\u0093 LOF" | tempusoncoprint$typeofvariant=="Missense variant GOF" | tempusoncoprint$typeofvariant=="Missense variant LOF" | tempusoncoprint$typeofvariant=="Missense variant -" | tempusoncoprint$typeofvariant=="Missense variant - GOF" | tempusoncoprint$typeofvariant=="Missense variant - LOF:" | tempusoncoprint$typeofvariant=="Missense variant- GOF" | tempusoncoprint$typeofvariant=="Missense variant LOF" | tempusoncoprint$typeofvariant=="Missense variant- LOF" | tempusoncoprint$typeofvariant=="Missense variant-LOF" | tempusoncoprint$typeofvariant=="Missense variant(exon 10) - GOF" | tempusoncoprint$typeofvariant=="Missense variant(exon 2) - GOF" | tempusoncoprint$typeofvariant=="Missense variant(exon 4) - GOF" | tempusoncoprint$typeofvariant=="Missense varian(exon 10) - GOF" | tempusoncoprint$typeofvariant=="Missense variant - LOF" | tempusoncoprint$typeofvariant=="Missesne variant" | tempusoncoprint$typeofvariant=="Missense variant -LOF", "MISSENSE", tempusoncoprint$variant)

tempusoncoprint$variant <- ifelse(tempusoncoprint$typeofvariant=="region variant" | tempusoncoprint$typeofvariant=="region variant- LOF" | tempusoncoprint$typeofvariant=="Splice" | tempusoncoprint$typeofvariant=="Splice region" | tempusoncoprint$typeofvariant=="Splice region vairant" | tempusoncoprint$typeofvariant=="Splice region vairant - LOF" | tempusoncoprint$typeofvariant=="splice region variant" | tempusoncoprint$typeofvariant=="Splice region variant" | tempusoncoprint$typeofvariant=="Splice region variant - LOF" | tempusoncoprint$typeofvariant=="Splice region variant -LOF" | tempusoncoprint$typeofvariant=="Splice region variant 창\u0080\u0093 GOF" | tempusoncoprint$typeofvariant=="Splice region variant 창\u0080\u0093 LOF" | tempusoncoprint$typeofvariant=="Splice region variant VAF" | tempusoncoprint$typeofvariant=="Splice region variant- GOF" | tempusoncoprint$typeofvariant=="Splice region variant- LOF" | tempusoncoprint$typeofvariant=="Splice region variant-(exon 3) 창\u0080\u0093 GOF", "PROMOTER", tempusoncoprint$variant)

tempusoncoprint$variant <- ifelse(tempusoncoprint$typeofvariant=="Start loss" | tempusoncoprint$typeofvariant=="Start Loss" | tempusoncoprint$typeofvariant=="StartStop gain loss- LOF" | tempusoncoprint$typeofvariant=="stop gain" | tempusoncoprint$typeofvariant=="Stop gain" | tempusoncoprint$typeofvariant=="Stop Gain" | tempusoncoprint$typeofvariant=="Stop gain - LOF" | tempusoncoprint$typeofvariant=="Stop gain LOF" | tempusoncoprint$typeofvariant=="Stop gain VAF" | tempusoncoprint$typeofvariant=="Stop gain-" | tempusoncoprint$typeofvariant=="Stop gain- GOF" | tempusoncoprint$typeofvariant=="Stop gain-LOF" | tempusoncoprint$typeofvariant=="Stop loss" | tempusoncoprint$typeofvariant=="Stop gain(exon 4) - LOF" | tempusoncoprint$typeofvariant=="Stopgain" | tempusoncoprint$typeofvariant=="Stop gain- LOF", "TRUNC", tempusoncoprint$variant)

tempusoncoprint$variant <- ifelse(is.na(tempusoncoprint$variant), "OTHER", tempusoncoprint$variant)

tempusoncoprint$alteration <- ifelse (is.na(tempusoncoprint$alteration), as.character(tempusoncoprint$variant_name_P), tempusoncoprint$alteration)

tempusoncoprint$alteration <- ifelse (tempusoncoprint$alteration=="", "MUT", tempusoncoprint$alteration)

tempusoncoprint <- within(tempusoncoprint, rm(StudyID, variant_name_P, typeofvariant)) #Removing alteration if not using CBioportal for Oncoprinter

write.table (tempusoncoprint, file="oncoprint.tsv", sep="\t", col.names=FALSE, row.names=FALSE)
```

```{r Ongoing mutational summary, echo=FALSE}
summary(tempusmergedupclean$TMB)
hist(tempusmergedupclean$TMB, main="Distribution of tumor mutational burden")
```
Only 88 of 139 reports contained percentile information; this was not included for analysis. Tumor mutational burden was measured in m/MB.

```{r Ongoing mutational summary-2, echo=FALSE}
kable(table(tempusmergedupclean$MicrosatelliteInstabilityStatu), caption="Distribution of MSI status", col.names=c("MSI status", "Freq"))
```
The majority of patients had stable MSI status in their tumors; only 1.5% showed MSI unstable disease. 

```{r Immune summary, echo=FALSE}
summary(tempusmergedupclean$b)
hist(tempusmergedupclean$TMB, main="Distribution of B cells in the tumor immune infiltrate")
summary(tempusmergedupclean$cd4)
hist(tempusmergedupclean$cd4, main="Distribution of CD4 cells in the tumor immune infiltrate")
summary(tempusmergedupclean$cd8)
hist(tempusmergedupclean$cd8, main="Distribution of CD8 cells in the tumor immune infiltrate")
summary(tempusmergedupclean$mac)
hist(tempusmergedupclean$mac, main="Distribution of macrophages in the tumor immune infiltrate")
summary(tempusmergedupclean$nk)
hist(tempusmergedupclean$nk, main="Distribution of NK cells in the tumor immune infiltrate")
summary(tempusmergedupclean$immune)
hist(tempusmergedupclean$immune, main="Immune scores as calculated by Tempus")
```
74 patients had "immune profile" data as calculated by Tempus. This represents a derivation from the RNA-seq data that was developed from a machine learning algorithm, NOT a pathology-based assessment. The immune score is a "support vector regression model for the immune cell estimates and formal thresholds for high or low are not in place as of yet."

#Recurrence and survival by ethnicity in our cohort
```{r Ethnicity recurrence and survival}
fitsurvival <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$vitalstatussurv)) ~tempusmergedupclean$bw, 
               data=tempusmergedupclean)
ggsurvplot(fitsurvival, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of patients by ethnicity")

fitrecurrence <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$recurrenceposneg)) ~tempusmergedupclean$bw, 
               data=tempusmergedupclean)
ggsurvplot(fitrecurrence, data=tempusmergedupclean, risk.table="percentage", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in patients by ethnicity")
```

#Mutations among different subgroups - Subtype and race (Not significant)
```{r Mutational summary among subgroups, warning=FALSE}
subtypeHERposHRneg <- tempusmergedupclean %>% filter(subtype=="HER+/HR-")
allgenesHERposHRneg <- colSums(!((subtypeHERposHRneg[,105:663])==""))
allgenesHERposHRneg["ERBB2/HER2"] <- sum(allgenesHERposHRneg["ERBB2"], allgenesHERposHRneg["ERBB2 (HER2)"], allgenesHERposHRneg["ERBB2(HER2)"], allgenesHERposHRneg["ERBB2HER2)"])
headHERposHRneg <- head(sort(allgenesHERposHRneg, decreasing=TRUE), 10)
headHERposHRneg

subtypeHERnegHRpos <- tempusmergedupclean %>% filter(subtype=="HER-/HR+")
allgenesHERnegHRpos <- colSums(!((subtypeHERnegHRpos[,105:663])==""))
allgenesHERnegHRpos["ERBB2/HER2"] <- sum(allgenesHERnegHRpos["ERBB2"], allgenesHERnegHRpos["ERBB2 (HER2)"], allgenesHERnegHRpos["ERBB2(HER2)"], allgenesHERnegHRpos["ERBB2HER2)"])
headHERnegHRpos <- head(sort(allgenesHERnegHRpos, decreasing=TRUE), 10)
headHERnegHRpos

subtypeTNBC <- tempusmergedupclean %>% filter(subtype=="TNBC")
allgenesTNBC <- colSums(!((subtypeTNBC[,105:663])==""))
allgenesTNBC["ERBB2/HER2"] <- sum(allgenesTNBC["ERBB2"], allgenesTNBC["ERBB2 (HER2)"], allgenesTNBC["ERBB2(HER2)"], allgenesTNBC["ERBB2HER2)"])
headTNBC<- head(sort(allgenesTNBC, decreasing=TRUE), 10)
headTNBC

subtypeHERposHRpos <- tempusmergedupclean %>% filter(subtype=="HER+/HR+")
allgenesHERposHRpos <- colSums(!((subtypeHERposHRpos[,105:663])==""))
allgenesHERposHRpos["ERBB2/HER2"] <- sum(allgenesHERposHRpos["ERBB2"], allgenesHERposHRpos["ERBB2 (HER2)"], allgenesHERposHRpos["ERBB2(HER2)"], allgenesHERposHRpos["ERBB2HER2)"])
headHERposHRpos <- head(sort(allgenesHERposHRpos, decreasing=TRUE), 10)
headHERposHRpos

chisq.test(headHERnegHRpos, headHERposHRneg, headTNBC, headHERposHRpos)

subtypewhite <- tempusmergedupclean %>% filter(race=="White")
allgeneswhite<- colSums(!((subtypewhite[,105:663])==""))
allgeneswhite["ERBB2/HER2"] <- sum(allgeneswhite["ERBB2"], allgeneswhite["ERBB2 (HER2)"], allgeneswhite["ERBB2(HER2)"], allgeneswhite["ERBB2HER2)"])
headwhite <- head(sort(allgeneswhite, decreasing=TRUE), 5)
headwhite

subtypeblack <- tempusmergedupclean %>% filter(race=="Black")
allgenesblack<- colSums(!((subtypeblack[,105:663])==""))
allgenesblack["ERBB2/HER2"] <- sum(allgenesblack["ERBB2"], allgenesblack["ERBB2 (HER2)"], allgenesblack["ERBB2(HER2)"], allgenesblack["ERBB2HER2)"])
headblack <- head(sort(allgenesblack, decreasing=TRUE), 5)
headblack

chisq.test(headwhite, headblack)
```
We did not see a significant difference in our cohort between breast cancer subtype or race and the most commonly mutated genes (outside of overrepresentation of ERBB2/HER2 in the HER2+ subtype).

```{r Outcomes by mutation status}
tempusmergedupclean$TP53YN <- ifelse (tempusmergedupclean$TP53=="", 0, 1)

fitsurvival <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$vitalstatussurv)) ~tempusmergedupclean$TP53YN, 
               data=tempusmergedupclean)
ggsurvplot(fitsurvival, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of patients by TP53 mutation status")

fitrecurrence <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$recurrenceposneg)) ~tempusmergedupclean$TP53YN, 
               data=tempusmergedupclean)
ggsurvplot(fitrecurrence, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in patients by TP53 mutation status")

tempusmergedupclean$BRCAYN <- ifelse (tempusmergedupclean$BRCA1=="" & tempusmergedupclean$BRCA2=="", 0, 1)

fitsurvival <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$vitalstatussurv)) ~tempusmergedupclean$BRCAYN, 
               data=tempusmergedupclean)
ggsurvplot(fitsurvival, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of patients by BRCA mutation status")

fitrecurrence <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$recurrenceposneg)) ~tempusmergedupclean$BRCAYN, 
               data=tempusmergedupclean)
ggsurvplot(fitrecurrence, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in patients by BRCA mutation status")

tempusmergedupclean$ARYN <- ifelse (tempusmergedupclean$AR=="", 0, 1)

fitsurvival <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$vitalstatussurv)) ~tempusmergedupclean$ARYN, 
               data=tempusmergedupclean)
ggsurvplot(fitsurvival, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of patients by AR mutation status")

fitrecurrence <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$recurrenceposneg)) ~tempusmergedupclean$ARYN, 
               data=tempusmergedupclean)
ggsurvplot(fitrecurrence, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in patients by AR mutation status")

tempusmergedupclean$PIK3CAYN <- ifelse (tempusmergedupclean$PIK3CA=="", 0, 1)

fitsurvival <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$vitalstatussurv)) ~tempusmergedupclean$PIK3CAYN, 
               data=tempusmergedupclean)
ggsurvplot(fitsurvival, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of patients by PIK3CA mutation status")

fitrecurrence <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$recurrenceposneg)) ~tempusmergedupclean$PIK3CAYN, 
               data=tempusmergedupclean)
ggsurvplot(fitrecurrence, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in patients by PIK3CA mutation status")
```

#Immune profile data and association with survival and recurrence
```{r Immune profile recurrence and survival}
tempusmergedupclean$TMBhighlow <- ifelse(tempusmergedupclean$TMB>20, 1, 0)
  
fitsurvival <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$vitalstatussurv)) ~tempusmergedupclean$TMBhighlow, 
               data=tempusmergedupclean)
ggsurvplot(fitsurvival, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of patients by TMB mutation status (high>20 m/MB)")

fitrecurrence <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$recurrenceposneg)) ~tempusmergedupclean$TMBhighlow, 
               data=tempusmergedupclean)
ggsurvplot(fitrecurrence, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in patients by TMB (high>20 m/mB")

tempusmergedupclean$immunemedian <- ifelse(tempusmergedupclean$immune>0.12, 1, 0)

fitsurvival <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$vitalstatussurv)) ~tempusmergedupclean$immunemedian, 
               data=tempusmergedupclean)
ggsurvplot(fitsurvival, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of patients by immune score greater or less than the median")

fitrecurrence <- survfit(Surv(time=tempusmergedupclean$timefu, as.numeric(tempusmergedupclean$recurrenceposneg)) ~tempusmergedupclean$immunemedian, 
               data=tempusmergedupclean)
ggsurvplot(fitrecurrence, data=tempusmergedupclean, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in patients by immune score greater or less than the median")

surv_object <- Surv(time=tempusmergedupclean$timefu, event=as.numeric(tempusmergedupclean$vitalstatussurv))
     
fit.coxph <- coxph(surv_object ~ tempusmergedupclean$TMBhighlow + tempusmergedupclean$immunemedian,
                   data = tempusmergedupclean)
                   
ggforest(fit.coxph, data = tempusmergedupclean, main="Comparative hazard ratios of immune measurements in patient tumors")

```

Based on Chalmers et al (https://genomemedicine.biomedcentral.com/track/pdf/10.1186/s13073-017-0424-2?site=genomemedicine.biomedcentral.com), we defined a high TMB as >20. We saw a nonsignificant trend towards increased mutational burden and immune infiltrate being associated with worse survival in these patients. 

```{r Neoadjuvant patients}
tempusneo <- tempusmergedupclean %>% filter(tempusmergedupclean$neoadj==1 & !(StudyID=="MB-0138") & !(StudyID=="MB-1020") & !(StudyID=="MB-1049"))

kable(table(tempusneo$recurrenceposneg), caption = "Neoadjuvant patients who experienced recurrence", col.names=c("Recurrence", "Freq"))
kable(table(tempusneo$vitalstatus), caption = "Neoadjuvant patients who did not survive (0=died)", col.names=c("Vital status", "Freq"))

fitsurvival <- survfit(Surv(time=tempusneo$timefu, as.numeric(tempusneo$vitalstatussurv)) ~tempusneo$TP53YN, 
               data=tempusneo)
ggsurvplot(fitsurvival, data=tempusneo, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, title="Survival of neoadjuvant patients by TP53 status")

fitrecurrence <- survfit(Surv(time=tempusneo$timefu, as.numeric(tempusneo$recurrenceposneg)) ~tempusneo$TP53YN, 
               data=tempusneo)
ggsurvplot(fitrecurrence, data=tempusneo, risk.table="absolute", pval=TRUE, xlab = "Time in years", xlim=c(0, 10), break.time.by=1, ylab="Recurrence", title="Likelihood of recurrence in neoadjuvant patients by TP53 mutation status")
```

#Heterogeneity between sentinel lymph nodes and primary tumors
```{r Lymph node differences relative to primary tumor}
tempusbreastlymph <- tempusmerge %>% filter((StudyID %in% duplicates$StudyID) & !(Tumorspecimen=="") & !(StudyID=="PN0263"))
  subtypelymph <- tempusbreastlymph %>% filter(Tumorspecimen=="Lymph node")
  subtypebreast <- tempusbreastlymph %>% filter(Tumorspecimen=="Breast")

  allgenesbreast<- colSums(!((subtypebreast[,105:663])==""))
  allgenesbreast["ERBB2/HER2"] <- sum(allgenesbreast["ERBB2"], allgenesbreast["ERBB2 (HER2)"], allgenesbreast["ERBB2(HER2)"], allgenesbreast["ERBB2HER2)"])
  print(head(sort(allgenesbreast, decreasing=TRUE)))
  
  allgeneslymph<- colSums(!((subtypelymph[,105:663])==""))
  allgeneslymph["ERBB2/HER2"] <- sum(allgeneslymph["ERBB2"], allgeneslymph["ERBB2 (HER2)"], allgeneslymph["ERBB2(HER2)"], allgeneslymph["ERBB2HER2)"])
  print(head(sort(allgeneslymph, decreasing=TRUE)))
  
```
We did see heterogeneity between the mutations represented in the lymph nodes and the mutations represented in the primary breast tumors for those samples. Two of the lymph nodes had 20% purity, 2 had 40% purity and 2 had 50% purity. 

#Extremely mutated patients x 2 - case examples
```{r Look at characteristics of extreme patients}
tempusmergedupclean$StudyID[tempusmergedupclean$genecount==max(tempusmergedupclean$genecount)]
```
The patient who had the most unique genes mutated in her tumor in our sample (with 184) was black, was living as of last f/u with iniital diagnosis in 2001 and f/u in 2011. HR+/HER2- disease without known recurrence. Her TMB was 71.3 and in the 100th percentile, MSI status was equivocal. The patient who had the highest number of variants (9) within a single gene was white, was dead with initial diagnosis in 2000 and f/u in 2006 with HR+, HER2 unknown disease with unknown recurrence status. Her TMB was 58.3 and in the 100th percentile; genecount was 3rd highest in the group, MSI status was normal. Both patients were cigarette smokers although cigarette smoking was not significantly associated with any of our mutation variables.  

```{r Descriptive statistics on patients with multiple variants}
sameptsamegenemultivariant <- tempusmutation %>% group_by (StudyID, Gene) %>% filter((duplicated(StudyID) & duplicated(Gene) & duplicated(AccessionNo)) | (duplicated(StudyID, fromLast=TRUE) & duplicated(AccessionNo, fromLast=TRUE) & duplicated(Gene, fromLast=TRUE))) %>% dplyr::select("StudyID", "AccessionNo", "Tumorspecimen", "Gene", "variant_name_C", "variant_name_P", "NM_number", "typeofvariant", "genecount", "variantcount")
```
